---
title: 'Time Series with R: Manipulating Time Series Data in R with xts & zoo'
author: "Timo Meiendresch"
date: "15 January 2019"
output: 
  html_document
  
---
```{r, echo=FALSE}
# Cleaning
rm(list=ls())
graphics.off()
```

## Ch.1 Introducing xts and zoo objects

An xts object consists of a matrix + index (time).

```{r, message=FALSE}
library(xts)
```

Step 1: Create a simple matrix and an index of two consecutive days. 
```{r}
x <- matrix(1:4, 2,2 )
x

idx <- as.Date(c("2019-01-01", "2019-02-01"))
idx
```

Step 2: Bring matrix and index together as an xts element using the `xts()`function. Take a look at the structure with `str(x)`.
```{r}
X <- xts(x, order.by=idx)
str(X)
X

```
The xts() constructor allows to specify some more useful options, especially to take care of the frequency and the time zone. Note that xts is a subclass of `zoo`.

#### Deconstructing xts
We can use `coredata()`to extract the data component from an `xts` element and `index()` to extract the index (times). 

## Importing, exporting and converting time series

Use `as.xts()` to convert other time series into xts. Here, `ts` to `xts`.
```{r}
# Load data from R datasets
data(sunspots)
class(sunspots)

sunspots_xts <- as.xts(sunspots)
head(sunspots_xts)
```

We can also apply the `as.xts` function directly when loading the data, i.e. as: 
```{r, eval=FALSE}
as.xts(read.table("file"))
as.xts(read.csv("file"))
```

## Ch. 2 Basic Manipulations

```{r}
data(edhec, package="PerformanceAnalytics")
head(edhec["2009", 1])

# Select only variables from Jan09 to Mar09
edhec["2009-01/03",1]

```

## Other extraction techniques
Furthermore, we can select rows as:

```{r, eval=FALSE}
x[c(1,2,3),]
x[index(x) > "2016-08-20"]
dates <- as.POSIXct(c("2016-06-25","2016-06-27"))
```

## Update and replace elements
```{r}
# take subset of edhec first
edhec_09 <- edhec["2009", 1]

# replace two random values of edhec_09 with NAs
edhec_09[sample(1:length(edhec_09), 2)] <- NA
edhec_09

``` 

## Methods to find periods in your data
For the next few methods we'll use the sample data from the xts package 

```{r}
data(sample_matrix, package = "xts")
class(sample_matrix)
OHLC <- as.xts(sample_matrix)
class(OHLC)
dim(OHLC)
head(OHLC, 3)
```
We can use the `first()`and `last()` function to get  the "first x days/months/etc." or the "last x days/months/etc.". 

```{r}
first(OHLC[,"Close"], "3 days")
head(last(OHLC[,"Close"], "1 month"),3)
```

We can also use integers to specify n 
```{r}
first(OHLC[,"Close"], n=3)
```

`first()` and `last()` can be nested for internal intervals 
```{r}
# take last 3 days of first 2 months
# Backwards induction
last(first(OHLC[,"Close"], "2 months"), "3 days")

# negative specifications
first(first(OHLC[, "Close"], "1 months"),"-27 days")
```

## Math operations using xts
xts is a matrix. Math operations are on the intersection of times (common dates). 
```{r}
# Take subset as first matrix
a <- first(OHLC[,"Close"], "3 days")
a
# Create a second xts matrix 
x <- matrix(rnorm(4,20,1))

# Note that idx is in the same format as "a"
idx <- seq(from = as.POSIXct("2007-01-03"), to = as.POSIXct("2007-01-06"), "1 day")
b <- xts(x, order.by=idx)
b
cbind(a, b)
```

Math operations are on the intersection of times only. 
```{r}
a + b # Applied only if we have the same dates

# add a to b using index of b
b + merge(a, index(b))

# fill missings with 0 <- takes values from b. 
b + merge(a, index(b), fill = 0)

# use locf to fill in missing values
b + merge(a, index(b), fill = na.locf)
```

## Ch. 3 Merging time series 
Series can be combined by column using `cbind()` or `merge()`. These joins on index (i.e. by time). Inner, outer, left and right joins are possible.

```{r}
# Outer join: all available dates
merge(a, b) # defaul join="outer"

# Inner join: Only common points in time
merge(a, b, join="inner")

# Right join: Joining according to index of right series dates
merge(a, b, join="right")

# Use fill to fill with a specific number
merge(a, b, join="right", fill=-1)

merge(a, c(1,1,1))
merge(a, 3)
merge(a, as.POSIXct("2019-01-01"))
```
Next: Consider rbind() to combine series by row. Rows are inserted in time order. All rows in rbind() must have a time and the numbers of columns must match.

```{r}
rbind(a, b)
```

## Handling missingness
locf means "las observation carried forward" and can be used to deal with NAs. 
```{r}
# Create a series with NAs
z <- rbind(a,b)
z[c(3,4)] <- NA
z

# use na.locf()
cbind(z, na.locf(z), na.locf(z, fromLast = TRUE))
```
In the last series last observation has been carried forward from before and from later observations. 

Other functions to deal with NAs:

* `na.fill(object, fill, ...)`
*  `na.trim(object, ...)`
* `na.omit(object, ...)`

We can also interpolate NAs with: 

* `na.approx(object, ...)`

## Lags and differences
Use `lag()` to shift observations in time. number of lags k and NA handling can be modified. 

```{r}
# compare standard differences to initial series
cbind(b, diff(b))

cbind(b, lag(b, k = -1))

# Create own first differences
z <- cbind(b, lag(b))
z$first_diff <- z$b - z$lag.b.
z

```

## Ch. 4 Apply functions by time
Two main approaches to apply functions on discrete periods or intervals. Using `period.apply(x, INDEX, FUN, ...)` or `split()`. 

To find the index of the last observation of an interval, we can use the function `endpoints(x, on="years")`.

```{r}
# Using period.apply(x, INDEX, FUN, ...)
ep <- endpoints(OHLC, "months", k = 2)
OHLC[ep]$Close

# k = 2 specifies that every two months are taken
period.apply(OHLC$Close, INDEX=ep, FUN=mean)
```
To reach a similar result we can also use shortcut functions: 

* `apply.monthly(), apply.yearly(), apply.quarterly(), etc.`

```{r}
apply.monthly(OHLC$Close, mean)

```

The `split(x, f="months")` function splits data into chunks of time lists. 
```{r}
OHLC_monthly <- split(OHLC[,"Close"], f="months")

lapply(X = OHLC_monthly, FUN=mean)
```

## Converting periodicity
Function `to.period()` to aggregate OHLC data.

```{r}
to.period(x = OHLC$Close, period = "months", name="OHLC")

# Alternatively use the shortcut function
to.monthly(OHLC$Close, name = "OHLC")
```
Note that `OHLC = FALSE` is an option and that we can

## Rolling functions
First, `split- lapply - do.call(rbind)` paradigm is useful to split a time series apply certain cumulative calculations before binding the time series back together. 
```{r}
OHLC_monthly <- split(OHLC[,1], f="months")
OHLC_monthly <- lapply(OHLC_monthly, FUN=cumsum)
OHLC_monthly <- do.call(rbind, OHLC_monthly)
```

Rolling windows to apply a function on a certain period, eg. always 3 days, etc. 
```{r}
# calculate three-day mean of OHLC
head(rollapply(OHLC[,4], 3, mean))
```

## Ch. 5 Index, Attributes, and TimeZones
```{r}
# Check index and timezone
index(OHLC)[1:3]
indexClass(OHLC)
indexTZ(OHLC)

# Change index format 
head(OHLC[,1],3)
indexFormat(OHLC) <- "%d %b %Y"
head(OHLC[,1],3)
```
Always check and set your time zone to avoid surprises. This can be done with the xts constructor as it has a `tzone=...` argument

## Periodicity
Identify periods of the data you are working with. Useful for regular timestamps

```{r}
# Check periodicity of OHLC
periodicity(OHLC)
periodicity(to.monthly(OHLC))

# counting days, months, seconds, etc.
nquarters(OHLC)
nmonths(OHLC)
nweeks(OHLC)

# weekly index
head(.indexwday(OHLC))

# check monthly index 
head(.indexmday(OHLC))

# check yearly index
tail(.indexyday(OHLC))

```

Left out: Modifying timestamps to remove times to allow for uniqueness (`make.index.unique()`) or, if too many time stamps are availabe the `align.time()`command will help, setting the n argument to the number of second you'd like to round to.  
