---
title: "GARCH models in R"
author: "Timo Meiendresch"
knit: (function(input_file, encoding) {
  out_dir <- 'html_outputs';
  rmarkdown::render(input_file,
  encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'garch models in R.html'))})
---

```{r, message=FALSE, echo=FALSE}
rm(list=ls())
graphics.off()
```

```{r, message=FALSE}
# libraries
library("PerformanceAnalytics")
library("xts")
library("quantmod")
library("ggplot2")
library("rugarch")
```

# 1. Standard GARCH models in R 

Starting with calculating relative financial gains and losses, expressed in terms of returns:

$$R_t = \frac{P_t - P_{t-1}}{P_{t-1}} $$

```{r, message=FALSE}
# Get SP500 data
getSymbols("^GSPC")

# Plot data
autoplot(GSPC$GSPC.Close)

# Calculate returns based on closing price
sp500c <- Cl(GSPC)
sp500ret <- na.omit(CalculateReturns(sp500c))

# Plot returns
autoplot(sp500ret) +
  ggtitle("Returns of S&P 500")+
  xlab("Year")+
  ylab("Return")
```

- Substantially higher volatilitily during financial crisis

Compute empirical standard deviation using sd(), which is the daily volatility when applied to returns. Corresponding formula for T daily returns: 
$$ \hat{\sigma} = \sqrt{ \frac{1}{T-1} \sum_{t=1}^T (R_t - \hat{\mu})^2  } $$
Annualized volatility $= \sqrt{252} \times $ daily volatility.


```{r}
# daily sd
sd(sp500ret)

# annualized volatility
sqrt(252) * sd(sp500ret)
```

### Rolling volatility estimation
- Window width: 22 (trading days) per month
- function: `chart.RollingPerformance()`

```{r}
chart.RollingPerformance(R = sp500ret,
                         width = 22, 
                         FUN = "sd.annualized",
                         scale= 252, 
                         main = "Rolling 1 month vola")

chart.RollingPerformance(R = sp500ret,
                         width = 22*6, 
                         FUN = "sd.annualized",
                         scale= 252, 
                         main = "Rolling 6 month vola")


```

### Garch models
- Estimation of $\sigma_t$ requires time series models, like GARCH 

Notation and steps of GARCH-Analysis: 

1. Input: Time series of returns
2. Prediction at time $t-1$ about $R_t$ based on the information set $I_{t-1}$. 
3. Predicting the mean return $\mu$. The objective is to minimize the prediction error $e_t = R_t - \mu$ 
$$\mu = E[R_t | I_{t-1}] $$
4.  We then predict the variance as

$$ 
\begin{align}
\sigma^2_t &= var(R_t | I_{t-1}) \\
 &= E[(R_t - \mu_t)^2 | I_{t-1}] \\
 &= E[e^2_t | I_{t-1}]
\end{align}
$$

In practice, we need an equation that maps past returns into a prediction of the mean

- Rolling mean model: $\mu_t = \frac{1}{M} \sum_{i=1}^{M} R_{t-1}$. Alternatively: AR/MA models 

We also need an equatoin that maps the past returns into predictions of the variance.

- Rolling variance model: $\sigma^2_t = \frac{1}{M} \sum_{t=1}^M e_{t-i}^2 $

More weights to recent observations:

- ARCH(p) model: $\sigma_t^2 = c + \sum_{i=1}^p \; \alpha_i e_{t-i}^2$

However, normally a GARCH(1,1) model is used

- GARCH(1,1) model: $\sigma_t^2 = c + \alpha \, e_{t-1}^2 + \beta \, \sigma_{t-1}^2$

It can be shown that the errors $e_t^2$ follow an ARMA representation. 

### Parameter Restrictions GARCH

1. c $\alpha$ and $\beta$ are $>0$ to ensure that variance is always positive.
2. $\alpha + \beta < 1$ to ensure that the predicted variance $\sigma_t^2$ returns to the long run variance (stationarity).

The variance returns to the mean and long-run variance equals: $$\frac{c}{1- \alpha - \beta}$$

```{r}
# set parameter values
alpha <- 0.1
beta <- 0.8
omega <- var(sp500ret) * (1-alpha - beta)

# set series of prediction errors
e <- sp500ret - mean(sp500ret) 
e2 <- e^2

## Predict variance for each observation
# prepare var to be filled in loop
n <- length(sp500ret)
predvar <- rep(NA, n)

# initialize at sample var
predvar[1] <- var(sp500ret)

# Loop 
for (t in 2:n){
  # GARCH (1,1):
  predvar[t] <- omega + alpha * e2[t-1] + beta * predvar[t-1] 
}

# volatility = sqrt(variance)
predvol <- sqrt(predvar)

# Create xts object 
predvol <- xts(predvol, order.by = time(sp500ret))

# Compare with unconditional volatility
uncvol <- sqrt(omega / (1 - alpha - beta))
uncvol <- xts(rep(uncvol, n), order.by = time(sp500ret))

plot(predvol)
lines(uncvol, col = "red", lwd=2)
```

```{r}
# Compute the mean daily return
m <- mean(sp500ret)

# Define the series of prediction errors
e <- sp500ret - m

# Plot the absolute value of the prediction errors
par(mfrow = c(2,1),mar = c(3, 2, 2, 2))
plot(abs(e))

# Plot the acf of the absolute prediction errors
acf(abs(e))
```

### The rugarch package
Standard GARCH(1,1) model:
- $R_t = \mu + e_t$
- $e_t \sim N(0, \sigma_t^2)$
- $\sigma_t^2 = \omega + \alpha \, e_{t-1}^2 + \beta \, \sigma_{t-1}^2$

Estimation by maximum likelihood: Find the parameter values for which the GARCH model is most likely to have generated the observed return series.

Workflow:

1. Specify GARCH model ($\mu, \sigma_t^2, e_t$) with `ugarchspec()`
2. Estimate the GARCH model of the time series with returns $R_1, ..., R_T$ using `ugarchfit()`
3. Use the estimated GARCH model to make volatility predctions for $R_{T+1}, ...$

```{r}
# Specify GARCH(1,1)
garchspec <- ugarchspec(
  mean.model = list(armaOrder = c(0,0)),
  variance.model = list(model = "sGARCH"),
  distribution.model = "norm")

# Estimate
garchfit <- ugarchfit(spec = garchspec,
                      data = sp500ret)

# Forecast 10 periods ahead
garchfc <- ugarchforecast(fitORspec = garchfit, n.ahead = 10)

# Extract specific values
garchcoef <- coef(garchfit)
garchuncvar <- uncvariance(garchfit)
garchmean <- fitted(garchfit)
garchvol <- sigma(garchfit)
```

The model is estimated as
$$ 
\begin{align}
R_t &= 7.1 \cdot 10^{-4} + e_t \\
e_t &\sim N(0, \hat{\sigma}_t^2) \\
\hat{\sigma}_t^2 &= 2.6 \cdot 10^{-6} + 0.13 \, e_{t-1}^2 + 0.85 \, \hat{\sigma}_{t-1}^2
\end{align}
$$

```{r}
# plot estimated volatility
garchvol <- sigma(garchfit)
plot(garchvol)

# forecasts only
sigma(garchfc)

# give mean of forecast for return
fitted(garchfc)

```

### Tactical Asset Allocation
- Invest percentage of w in a risky asset with volatility $\sigma_t$ and keep (1-w) on a risk-free bank deposit account, yielding the volatility
$$\sigma_p = w \sigma_t$$
- How to set w? Volatility Targeting! Solve for w and substitute desired volatility $\sigma_p$, as well as estimated volatility $\sigma_t$

$$w = \frac{\sigma_p}{\sigma_t}$$

```{r}
annualvol <- sqrt(252) * sigma(garchfit)

# weight risky asset
vt_weights <- 0.05 /annualvol

# plot weights and annualvol
plot(merge(annualvol, vt_weights), multi.panel = TRUE)
```

The plot shows the weight w, which ought to be invested in the SP500 if the desired volatility of the two-asset portfolio is 5%. 

# 2. Improvements of the normal GARCH model

- tbc
